
---
title: "Machine Learning Project MJ"
author: "Mazen Jabbour"
date: "11/22/2020"
output: html_document
---

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 

#Loading R packages

```{r, echo=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)

```

#Loading Data

```{r, echo=FALSE}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}

```

#Data as CSV

```{r, echo=FALSE}
trainRaw <- read.csv("./data/pml-training.csv")
testRaw <- read.csv("./data/pml-testing.csv")
dim(trainRaw)
dim(testRaw)
```

The training data set contains 19622 observations and 160 variables.
The testing data set contains 20 observations and 160 variables.
Predictor variable: classe

#Data Cleaning
#This step will include: missing value analysis, and keep only the variables we need

```{r, echo=FALSE}
sum(complete.cases(trainRaw))
```

#Removing missing values

```{r, echo=FALSE}
trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0] 
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0] 
```

#Keeping only the necessary variables for the analysis:

```{r, echo=FALSE}
classe <- trainRaw$classe
trainRemove <- grepl("^X|timestamp|window", names(trainRaw))
trainRaw <- trainRaw[, !trainRemove]
trainCleaned <- trainRaw[, sapply(trainRaw, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testRaw))
testRaw <- testRaw[, !testRemove]
testCleaned <- testRaw[, sapply(testRaw, is.numeric)]
dim(trainRaw)
dim(testRaw)
```

The adjusted training data set contains 19622 observations and 87 variables.
The adjusted testing data set contains 20 observations and 54 variables.

#Data Slicing

#We will split the cleaned training set into a pure training data set (70%) and a validation data set (30%). 

```{r, echo=FALSE}
set.seed(22519) # For reproducibile purpose
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]

```

#The model

We fit a predictive model for activity recognition using Random Forest algorithm. 
We will use 5-fold cross validation when applying the algorithm.

```{r, echo=FALSE}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf

```

#Estimating the performance of the model on the validation data set.

```{r, echo=FALSE}
predictRf <- predict(modelRf, testData)
confusionMatrix(table(testData$classe, predictRf)) 

```


The accuracy of the model is 99.42% and the estimated out-of-sample error is 0.58%.

##Graphs

#Correlation Matrix
```{r, echo=FALSE}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```

#Decision Tree

```{r, echo=FALSE}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) # fast plot

```

